---
title: "Data607_HW3"
author: "Sean Connin"
date: "02/19/2021"
output:
  html_document:
    df_print: paged
    theme: lumen
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

#open libraries

library(stringr)
library(magrittr)
library(tidyverse)
library(readr)
library(kableExtra)
```

## 1. College Majors Data Set

Using the 173 majors listed in fivethirtyeight.com’s College Majors dataset [https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/], provide code that identifies the majors that contain either "DATA" or "STATISTICS"

*Approach:*

1. Import dataset
2. Use str_detect() to identify strings containing "STATISTICS" or "DATA"
3. Subset rows with filter() retrieve these strings


```{r}


majors <- read_csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/majors-list.csv")

majors%<>%filter(str_detect(Major, "STATISTICS|DATA"))

majors%>%kbl()%>%kable_material(c("striped"))
                           
```
## 2. Write code that transforms the data below:

[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"
[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  
[9] "elderberry"   "lime"         "lychee"       "mulberry"    
[13] "olive"        "salal berry"

Into a format like this:

c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")

*Approach: adapted from https://bit.ly/2N4Q6zW*

1. shQuote() - dbl quote a string to be passed into os shell, cmd is Windows default
2. paste() - create a character string and separate results with a comma  via the collapse argument
3. cat() - outputs as a character vector
```{r}
d<- c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")


d<-(cat(paste(shQuote(d, type="cmd"), collapse=", ")))




```
## 3. Describe, in words, what these expressions will match:

(.)\1\1  - <span style="color: green;">*missing quotes, no match, \ treated as a literal*</span>

"(.)(.)\\2\\1" - <span style="color: green;">*matches two single character capture groups followed by their reverse order.*</span>

<span style="color: green;">*For example: if x= "bannnaannana" the match is "naan"*</span>

(..)\1 - <span style="color: green;">*missing quotes, no match, \ treated as a literal*</span>

"(.).\\1.\\1" - <span style="color: green;">*matches a single character capture group followed by another character, this pattern repeats, and ends with the first capture group.*</span>

 <span style="color: green;">*For example: if x= "bannnaannantnan" the match is "nantn"*</span>

*"(.)(.)(.).*\\3\\2\\1" - <span style="color: green;">*matches a sequence of three single character capture groups followed by a character that repeats one or more time followed by a sequence of the three capture groups in reverse order.*</span>

<span style="color: green;">*For example: if x= "banntaannantnan" the match is"anntaanna"*</span>

## 4. Construct regular expressions to match words that:

Start and end with the same character.

<span style="color: blue;">^(.).*\\1$</span>

Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)

<span style="color: blue;">(\\w{2}).*?(\\1)</span> Adapted from: https://bit.ly/2Ng5lps

Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.

<span style="color: blue;">(.).*?(\\1).*?(\\1).*?$</span>

